---
title: "R Notebook"
output: html_notebook
---

Load the packages
```{r}
install.packages('tidyverse')
install.packages('skimr')
library(tidyverse) #wrangle data
library(dplyr) #clean data
library(lubridate)  #wrangle date attributes
library(skimr) #get summary data
library(ggplot2) #visualize data
library(readr)
```

Prepare the data
```{r}
setwd("C:/Users/Emily/Downloads/data")

#Combine the data from Sep 2020 to Aug 2021 into one data frame.
bike_data <- rbind(
  read_csv("202009-divvy-tripdata.csv"), 
  read_csv("202010-divvy-tripdata.csv"),
  read_csv("202011-divvy-tripdata.csv"),
  read_csv("202012-divvy-tripdata.csv"),
  read_csv("202101-divvy-tripdata.csv"),
  read_csv("202102-divvy-tripdata.csv"),
  read_csv("202103-divvy-tripdata.csv"),
  read_csv("202104-divvy-tripdata.csv"),
  read_csv("202105-divvy-tripdata.csv"),
  read_csv("202106-divvy-tripdata.csv"),
  read_csv("202107-divvy-tripdata.csv"),
  read_csv("202108-divvy-tripdata.csv"))
```

Examine the combined data frame. 
```{r}
head(bike_data)
dim(bike_data)
colnames(bike_data)
summary(bike_data)

total_riders <- data.frame(table(bike_data$member_casual))
total_riders
total_types <- data.frame(table(bike_data$rideable_type))
total_types

write_csv(total_riders, "total_riders.csv")
write_csv(total_types, "total_types.csv")
```

Check for NA values
```{r}
sum(is.na(bike_data))
```

Drop columns we don't need: start_lat, start_lng, end_lat, end_lng
```{r}
bike_data <- bike_data %>% select(-c(start_lat, start_lng, end_lat, end_lng))
colnames(bike_data)
head(bike_data)
```


Add column “ride_length", which is the length of each ride from ended-at minus started_at, and format as HH:MM:SS.
Add column “day_of_week,”, and calculate the day of the week that each ride started.
```{r}
#bike_data$date <- as.Date(bike_data$started_at) #default format is yyyy-mm-dd
#bike_data$month <-  format(as.Date(bike_data$date), "%m")
#bike_data$day <- format(as.Date(bike_data$date), "%d")
#bike_data$year <- format(as.Date(bike_data$date), "%Y")

bike_data <- bike_data %>% mutate(ride_length = ended_at - started_at) %>% mutate(day_of_week = weekdays(as.Date(bike_data$started_at)))
head(bike_data)

#Convert ride_length from time to seconds (integer)
bike_data$ride_length <- as.numeric(bike_data$ride_length)
bike_data$ride_length <- as.numeric(bike_data$ride_length/60)
head(bike_data)

```

Removing the bad data and do analysis on the general ride length
```{r}
#check for data with negative ride length
bike_data <- bike_data[bike_data$ride_length>0,]

#check for data with ride length  more than 1 day (86400 seconds). There is 3484 rides that the rider use the bike for more than 1 day
sum(bike_data$ride_length > 1440)

#check for extreme outliner: 3356649 seconds or 38.85 days for max ride and 1 second for the least ride length
max(bike_data$ride_length)
min(bike_data$ride_length)

#check for mean and median
mean(bike_data$ride_length)
median(bike_data$ride_length)

#using summary to check min, max, median and mean
summary(bike_data$ride_length)
```
Aggregate to analyze the data based on user type: member vs casual

```{r}
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = mean)
aggregate(bike_data$ride_length ~ bike_data$member_casual, FUN = median)

#average ride time by each day for members vs casual
bike_data$day_of_week <- ordered(bike_data$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(bike_data$ride_length ~ bike_data$member_casual + bike_data$day_of_week, FUN = mean)
```
Further analysis into the stations
```{r}
head(count(bike_data, start_station_name, member_casual,  rideable_type, sort= TRUE))
head(count(bike_data, end_station_name, member_casual,  rideable_type, sort= TRUE))
```
Analyze ridership data by type and weekday
```{r}
bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```
Visualize the number of rides by rider type AND average duration
```{r}
par(mfrow=c(2,2))

bike_data %>%
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration=mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual))+
  geom_col(position="dodge")

bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")


```
Member vs Casual on short ride (less and one hour)
```{r}
one_hour_data <- bike_data %>% filter(ride_length < 60)
one_hour_data$ride_length_by_mins <- cut(one_hour_data$ride_length,breaks = 20)

ggplot(data = one_hour_data) + 
  geom_bar(mapping = aes(x = ride_length_by_mins, fill = member_casual)) +
  labs(title = "One hour ride length") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_x_discrete(labels=c("3", "6", "9", "12", "15", "18", "21", "24", "27", "30", "33", "36", "39", "42", "45", "48", "51", "54", "57", "60"))
```
Day of the week riding choices between member or causal
```{r}
ggplot(data = bike_data) +
  geom_bar(mapping = aes(x = factor(day_of_week), fill = rideable_type))+
  facet_wrap(~member_casual) +
  labs(title='riding choice during day of the week', x= 'day of the week' )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Check for peak time for bike usage between member vs casual
```{r}
hour_data <- bike_data
hour_data$start_hour <- as.numeric(format(strptime(bike_data$started_at,"%Y-%m-%d %H:%M:%OS"),'%H'))

ggplot(data = hour_data) + 
  geom_bar(mapping = aes(x = start_hour, fill = member_casual), stat = 'count') + 
  facet_wrap(~factor(day_of_week)) +
  labs(title = "bike usage by starting hour", x = "starting hour") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
  
```
Save as csv for further analysis
```{r}
write_csv(bike_data, "bikedata.csv")

#total and average weekly rides by rider type
summary_ride_weekly <- bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  
  
write_csv(summary_ride_weekly, "summary_ride_weekly.csv")

#total and average weekly rides by rider type
summary_ride_weekly_type <- bike_data %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday, rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  
  
write_csv(summary_ride_weekly_type, "summary_ride_weekly_type.csv")
  
#total and avg monthly rides by rider type
summary_month <- bike_data %>% 
  mutate(month = month(started_at, label = TRUE)) %>%  
  group_by(month,member_casual) %>%  
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%    
  arrange(month, member_casual)
write_csv(summary_month, "summary_ride_monthly.csv")


#most popular stations
popular_stations <- bike_data %>%
  mutate(station = start_station_name) %>%
  drop_na(start_station_name) %>%
  group_by(start_station_name, member_casual) %>%
  summarise(number_of_rides=n())

write_csv(popular_stations, "popular_stations.csv")
```


